Abstraction and Invariance for Algebraically Indexed Types

### 2012-07-10 paper/section-info-flow.tex

As described by Sabelfeld and Sands \cite{sabelfeld01per}, information
flow can be captured semantically by partial equivalence relations
(PERs). Abadi, Banerjee, Heintze and Riecke \cite{abadi99core} built a
\emph{Core Calculus for Dependency}, using a type system based around
a security level indexed monad $T_lA$, using PERs to prove the
information flow properties. Tse and Zdancewic \cite{tse04translating}
translated Abadi et al.'s calculus into System F, translating the
monadic type $T_lA$ to $\alpha_l \to A$ for some free type variable
$\alpha_l$, and using Reynolds' abstraction theorem to prove
information flow properties. For example, if the type variable
$\alpha_H$ represents high-level information, then the
non-interference property of the function $f$ could be expressed by
the System F type $A \to (\alpha_H \to B) \to C$. If a program cannot
generate a value of type $\alpha_H$, then it cannot access the value
of $B$, and hence is insensitive to the actual value. Relationships
between security levels are captured by postulating functions
$\alpha_{l_1} \to \alpha_{l_2}$ whenever $l_1$ is a lower security
level than $l_2$.

Using algebraic indexing, we refine Tse and Zdancewic's translation by
replacing each type variable $\alpha_l$ with a primitive type
$\tyPrim{T}{e}$ of representations of the truth of a logical
proposition $e$. The relationships between security levels are now
replaced by logical entailment, so we only have functions of type
$\tyPrim{T}{\phi} \to \tyPrim{T}{\psi}$ when $\phi$ entails
$\psi$. %Inspired by the PER interpretation of information flow, 
We instantiate our general relational framework to interpret
$\tyPrim{T}{\phi}$ as the identity relation if $\phi$ is true and the
empty relation if $\phi$ is false. This recovers the information flow
properties of Abadi {\em et al.} and Tse and Zdancewic.

@@

We have a single primitive type $\tyPrimNm{T}$, with
$\primTyArity(\tyPrimNm{T}) = [\mathsf{prop}]$ and index-erasure
semantics $\tyPrimSem{\tyPrimNm{X}} = \{ * \}$. Thus values of type
$\tyPrim{T}{\phi}$ have no run-time content; their only meaning is
given by the relational semantics. For the relational environments
$\relEnv{E}_\infoflow$, we use the compositional construction in
\autoref{sec:constr-rel-env}, with an arbitrary boolean algebra $L$
and $R_{\tyPrimNm{T}}(x) = \{ (*,*) \sepbar x = \top \}$.  The
primitive operations $\Gamma_\infoflow$ reflect logical consequence:
\begin{displaymath}
  \begin{array}{l}
  \begin{array}{@{}c@{\hspace{2em}}c}
    \mathrm{truth} : \tyPrim{T}{\top} &
    \mathrm{and}   : \forall p, q\mathord:\mathsf{prop}.\ \tyPrim{T}{p} \to \tyPrim{T}{q} \to \tyPrim{T}{p \land q}
  \end{array} \\
  \mathrm{up} : \forall p, q\mathord:\mathsf{prop}.\ \tyPrim{T}{p \land q} \to \tyPrim{T}{p}
\end{array}
\end{displaymath}
The combination of the \TirName{TyEq} rule and the primitive
$\mathrm{up}$ operation allow for logical entailment to be reflected
in programs: if we have $\phi \leq \psi$ and $M : \tyPrim{T}{\phi}$
then $\mathrm{up}\ M : \tyPrim{T}{\psi}$. Each of the primitive
operations has a trivial interpretation, due the index-erasure
interpretation of $\tyPrim{T}{\phi}$ as a one-element set, giving an
environment $\eta_\infoflow \in \ctxtSem{\Gamma}$. Less trivially, we
have this lemma:
\begin{lemma}\label{lem:environments-information-flow}
  For all $\Delta$ and $\rho \in \relEnv{E}_\infoflow$, $(\eta_\infoflow, \eta_\infoflow) \in \rsem{\Gamma_\infoflow}{\relEnv{E}_\infoflow}{\rho}$.
\end{lemma}

\paragraph{Information Flow} We think of logical expressions as
``composite principals''. That is, propositional variables represent
atomic principals that are combined using the logical connectives. We
interpret ``truth'' for principals as stating that a principal is true
when satisfied with the current state of affairs. Thus a relationship
$\phi \leq \psi$ indicates that satisfaction of the composite
principal $\phi$ implies satisfaction of the composite principal
$\psi$. In terms of security levels, the ordering is reversed: if a
high security principal is satisfied, then all of their underlings
must also be satisfied.

We adapt Tse and Zdancewic's translation of Abadi et al.'s monadic
type to our setting. We define a type abbreviation $T_\phi A =
\tyPrim{T}{\phi} \to A$, where $A$ is a type and $\phi$ is an
expression of sort $\mathsf{prop}$. For every $\phi$, we can endow the
types $T_\phi-$ with the structure of a monad. This is due to the fact
that it is an instance of the ``environment'' (or ``reader'') monad
\cite{jones95functional}. We read the types $T_\phi A$ as data of type
$A$ ``protected'' by the principal $\phi$.

As a consequence of the relational interpretation given above, it
follows that if we have an index expression $\phi$ that is interpreted
as some value other than $\top$ in a relational environment $\rho$,
then for all $x,x' \in \tySem{T_\phi\tyPrimNm{bool}}$, we have $(x,x')
\in \rsem{T_\phi\tyPrimNm{bool}}{\relEnv{E}_\infoflow}{\rho}$. Thus if
a principal is dissatisfied (i.e.,~$\phi \not= \top$), then data
protected by this principal is indistinguishable from any other
data. Thus a program cannot get access to the exact value. From this
and \thmref{thm:abstraction}, we obtain the following information flow
result:

\begin{theorem}
  Let $\phi$ and $\psi$ be index expressions of sort $\mathsf{prop}$
  in some indexing context $\Delta$, such that $\psi \not\leq
  \phi$. Then for all programs
  \begin{displaymath}
    \Delta; \Gamma_\infoflow,\Gamma \vdash M : T_\phi\tyPrimNm{bool} \to T_\psi\tyPrimNm{bool}
  \end{displaymath}
  and all terms $N_1,N_2$ of type $T_\phi\tyPrimNm{bool}$, $M\ N_1
  \stackrel{ctx}\approx M\ N_2$.  Thus there is no information flow
  from $M$'s input to its output.
\end{theorem}
This result also holds if we move to logics other than boolean
logic. For example, if our equational theory models intuitionistic
logic by taking the axioms of Heyting algebras, then the same non-flow
property for programs of type $T_{p\lor\lnot p}\tyPrimNm{bool} \to
T_\top\tyPrimNm{bool}$ holds, due to the lack of excluded middle. If
we take linear logic, then programs of type $T_{p\otimes
  p}\tyPrimNm{bool} \to T_p\tyPrimNm{bool}$ have no information flow
from their input, due to non-provability of $p \vdash p \otimes p$.

>>>>

As described by Sabelfeld and Sands \cite{sabelfeld01per}, information
flow can be captured semantically by partial equivalence relations
(PERs). Abadi, Banerjee, Heintze and Riecke \cite{abadi99core} built a
\emph{Core Calculus for Dependency}, using a type system based around
a security level indexed monad $T_lA$, using PERs to prove the
information flow properties. Tse and Zdancewic \cite{tse04translating}
translated Abadi \emph{et al.}'s calculus into System F, translating the
monadic type $T_lA$ to $\alpha_l \to A$ for some free type variable
$\alpha_l$, and using Reynolds' abstraction theorem to prove
information flow properties. For example, if the type variable
$\alpha_H$ represents high-level information, then the
non-interference property of the function $f$ could be expressed by
the System F type $A \to (\alpha_H \to B) \to C$. If a program cannot
generate a value of type $\alpha_H$, then it cannot access the value
of $B$, and hence is insensitive to the actual value. Relationships
between security levels are captured by postulating functions
$\alpha_{l_1} \to \alpha_{l_2}$ whenever $l_1$ is a lower security
level than $l_2$.

Using algebraic indexing, we refine Tse and Zdancewic's translation by
replacing each type variable $\alpha_l$ with a primitive type
$\tyPrim{T}{\phi}$ of representations of the truth of a logical
proposition $\phi$. The relationships between security levels are now
replaced by logical entailment, so we only have functions of type
$\tyPrim{T}{\phi} \to \tyPrim{T}{\psi}$ when $\phi$ entails
$\psi$. %Inspired by the PER interpretation of information flow,
We instantiate our general relational framework to interpret
$\tyPrim{T}{\phi}$ as the identity relation if $\phi$ is true and the
empty relation if $\phi$ is false. We shall see that this recovers the
information flow properties of Abadi {\em et al.} and Tse and
Zdancewic.

@@

We have a single primitive type $\tyPrimNm{T}$, with
$\primTyArity(\tyPrimNm{T}) = [\mathsf{prop}]$ and index-erasure
semantics $\tyPrimSem{\tyPrimNm{X}} = \{ * \}$. Thus values of type
$\tyPrim{T}{\phi}$ have no run-time content; their only meaning is
given by the relational semantics. For the relational environments
$\relEnv{E}_\infoflow$, we use the model-based construction in
\autoref{sec:constr-rel-env}, with an arbitrary boolean algebra $L$
and $R_{\tyPrimNm{T}}(x) = \{ (*,*) \sepbar x = \top \}$.  The
primitive operations $\Gamma_\infoflow$ reflect logical consequence:
\begin{displaymath}
  \begin{array}{l}
  \begin{array}{@{}c@{\hspace{2em}}c}
    \mathrm{truth} : \tyPrim{T}{\top} &
    \mathrm{and}   : \forall p, q\mathord:\mathsf{prop}.\ \tyPrim{T}{p} \to \tyPrim{T}{q} \to \tyPrim{T}{p \land q}
  \end{array} \\
  \mathrm{up} : \forall p, q\mathord:\mathsf{prop}.\ \tyPrim{T}{p \land q} \to \tyPrim{T}{p}
\end{array}
\end{displaymath}
The combination of the \TirName{TyEq} rule and the primitive
$\mathrm{up}$ operation allow for logical entailment to be reflected
in programs: if we have $\phi \leq \psi$ and $M : \tyPrim{T}{\phi}$
then $\mathrm{up}\ M : \tyPrim{T}{\psi}$. Each of the primitive
operations has a trivial interpretation, due the index-erasure
interpretation of $\tyPrim{T}{\phi}$ as a one-element set, giving an
environment $\eta_\infoflow \in \ctxtSem{\Gamma}$. Less trivially, we
have this lemma:
\begin{lemma}\label{lem:environments-information-flow}
  For all $\Delta$ and $\rho \in \relEnv{E}_\infoflow$, $(\eta_\infoflow, \eta_\infoflow) \in \rsem{\Gamma_\infoflow}{\relEnv{E}_\infoflow}{\rho}$.
\end{lemma}

\paragraph{Information Flow} We think of logical expressions as
``composite principals''. That is, propositional variables representing
atomic principals that are combined using the logical connectives. We
interpret ``truth'' for principals as stating that a principal is true
when satisfied with the current state of affairs. Thus a relationship
$\phi \leq \psi$ indicates that satisfaction of the composite
principal $\phi$ implies satisfaction of the composite principal
$\psi$. In terms of security levels, the ordering is reversed: if a
high security principal is satisfied, then all of their subordinates
must also be satisfied.

We adapt Tse and Zdancewic's translation of Abadi \emph{et al.}'s monadic
type to our setting. We define a type abbreviation $T_\phi A =
\tyPrim{T}{\phi} \to A$, where $A$ is a type and $\phi$ is an
expression of sort $\mathsf{prop}$. For every $\phi$, we can endow the
types $T_\phi-$ with the structure of a monad. This is due to the fact
that it is an instance of the ``environment'' (or ``reader'') monad
\cite{jones95functional}. We read the types $T_\phi A$ as data of type
$A$ ``protected'' by the principal $\phi$.

As a consequence of the relational interpretation given above, it
follows that if we have an index expression $\phi$ that is interpreted
as some value other than $\top$ in a relational environment $\rho$,
then for all $x,x' \in \tySem{T_\phi\tyPrimNm{bool}}$, we have $(x,x')
\in \rsem{T_\phi\tyPrimNm{bool}}{\relEnv{E}_\infoflow}{\rho}$. Thus if
a principal is dissatisfied (i.e.,~$\phi \not= \top$), then data
protected by this principal is indistinguishable from any other
data, and a program cannot get access to the exact value. From this
and \thmref{thm:abstraction}, we obtain the following information flow
result:

\begin{theorem}
  Let $\phi$ and $\psi$ be index expressions of sort $\mathsf{prop}$
  in some indexing context $\Delta$, such that $\psi \not\leq
  \phi$. Then for all programs
  \begin{displaymath}
    \Delta; \Gamma_\infoflow,\Gamma \vdash M : T_\phi\tyPrimNm{bool} \to T_\psi\tyPrimNm{bool}
  \end{displaymath}
  and all terms $N_1,N_2$ of type $T_\phi\tyPrimNm{bool}$, $M\ N_1
  \stackrel{ctx}\approx M\ N_2$.  Thus there is no information flow
  from $M$'s input to its output.
\end{theorem}
Note that if $\psi \leq \phi$, then it is always possible to write the
identity function with this type, using the $\mathrm{up}$ operation.
The theorem also holds if we move to logics other than boolean
logic. For example, if our equational theory models intuitionistic
logic by taking the axioms of Heyting algebras, then the same non-flow
property for programs of type $T_{p\lor\lnot p}\tyPrimNm{bool} \to
T_\top\tyPrimNm{bool}$ holds, due to the lack of excluded middle. If
we take linear logic, then programs of type $T_{p\otimes
  p}\tyPrimNm{bool} \to T_p\tyPrimNm{bool}$ have no information flow
from their input, due to non-provability of $p \vdash p \otimes p$.

### 2012-07-10 paper/section-instantiations.tex

We define the model of the indexing theory for the two-dimensional
geometry instantiation as follows. Each of the sorts is interpreted just as
its semantic counterpart: $\semSort{\SynTransl{2}} = \Transl{2}$, $
\semSort{\SynGL{2}} = \GL{2}$, $ \semSort{\SynOrth{2}} = \Orth{2}$,
and $ \semSort{\SynGL{1}} = \GL{1}$.
%  \begin{displaymath}
%    \begin{array}{@{}llll}
%      \semSort{\SynTransl{2}} = \Transl{2}
%      &
%      \semSort{\SynGL{2}} = \GL{2}
%      &
%      \semSort{\SynOrth{2}} = \Orth{2}
%      &
%      \semSort{\SynGL{1}} = \GL{1}
%    \end{array}
%  \end{displaymath}
Each of the index operations (e.g.,~the group structure and
determinant) is interpreted by the intended semantic operation, and
clearly satisfy the axioms in \exref{ex:two-dim-geo-axioms}. For the
relational interpretations of $\tyPrim{vec}{B,t}$ and
$\tyPrim{real}{s}$, we use the compositional construction from
\autoref{sec:constr-rel-env}, with $R_{\tyPrimNm{vec}}(B,\vec{t}) = \{
(\vec{v}, B\vec{v} + \vec{t}) \sepbar \vec{v} \in \mathbb{R}^2 \}$ and
$R_{\tyPrimNm{real}}(k) = \{ (x, kx) \sepbar x \in \mathbb{R} \}$. The
types of the primitive vector operations presented in
\autoref{sec:motivating-examples} are collected into a context
$\Gamma_\Geom$ with a straightforward interpretation $\eta_\Geom \in
\ctxtSem{\Gamma_\Geom}$.
  % \begin{displaymath}
  %   \begin{array}{l}
  %     R_{\tyPrimNm{vec}}(B,\vec{t}) = \{ (\vec{v}, B\vec{v} + \vec{t}) \sepbar \vec{v} \in \mathbb{R}^2 \} \\
  %     R_{\tyPrimNm{real}}(k) = \{ (x, kx) \sepbar x \in \mathbb{R} \}
  %   \end{array}
  % \end{displaymath}
%   The definitions of $R_{\tyPrimNm{vec}}$ and $R_{\tyPrimNm{real}}$
%   %are as expected. % given the informal definitions we gave in
% formalise the informal ones from~\autoref{sec:motivating-examples}, and 
%   the definitions of %We have set
%   $R^\bullet_{\tyPrimNm{vec}}$ and $R^\bullet_{\tyPrimNm{real}}$ 
%   %to relate $\vec{0}$ and $0$ to themselves to 
%   account for the
%   polymorphic zero constants we have assumed.
%\end{example}

@@

Consider the type of the area %of a triangle 
function from
\exref{ex:area-of-triangle-1}:
\begin{displaymath}
  \begin{array}{@{}l}
    \mathrm{area} : \forall B\mathord:\SynGL{2}, t\mathord:\SynTransl{2}.\ \\
    \hspace{0.8cm}\tyPrim{vec}{B, t} \to \tyPrim{vec}{B, t} \to \tyPrim{vec}{B, t} \to \tyPrim{real}{\abs{\det B}}
  \end{array}
\end{displaymath}
By \thmref{thm:abstraction}, we can derive the following free theorem. For all $B \in \GL{2}$, $\vec{t} \in \Transl{2}$, and $\vec{x}, \vec{y}, \vec{z} \in \mathbb{R}^2$, we have 
\begin{displaymath}
  \abs{\det B}(\tmSem{\mathrm{area}}\ \vec{x}\ \vec{y}\ \vec{z}) = \tmSem{\mathrm{area}}\ (B\vec{x} + \vec{t})\ (B\vec{y} + \vec{t})\ (B\vec{z} + \vec{t})
\end{displaymath}
Thus, directly from the type of the $\mathrm{area}$ function, we can
see that its index-erasure semantics is (a) invariant under
translations, and (b) if the inputs are subjected to a linear
transformation $B$, the output varies with the absolute value of the
determinant of $B$.

@@

For proving non-definability results, we consider the general case of
a sort $\mathsf{G}$ with the equational theory of abelian groups.  We
assume a group-indexed primitive type $\tyPrimNm{G}$ and set of
primitive operations typed by $\Gamma_\Grp$ that reflect the group
structure at the value level: identity $1:\tyPrim{G}{1}$, operation
$(\cdot): \forall a,b\mathord:\mathsf{G}.\ \tyPrim{G}{a} \to
\tyPrim{G}{b} \to \tyPrim{G}{a\cdot b}$ and inverse $\ ^{-1}: \forall
a\mathord:\mathsf{G}.\ \tyPrim{G}{a} \to \tyPrim{G}{a^{-1}}$. For the
index-erasure semantics, we assume some (not-necessarily abelian)
group $G$ and $\tyPrimSem{\tyPrimNm{G}} = G$.  The interpretation
$\eta_\Grp$ of the primitive operations in $\Gamma_\Grp$ is just the
unit, multiplication and inverse operations of $G$. One instance is the
sort~$\SynTransl{2}$ with $0$, $(+)$ and $\mathrm{negate}$ on
$\tyPrim{vec}{t}$.

We use the construction of non-compositional relational environments
from \autoref{sec:constr-rel-env}, with $R^\bullet_{\tyPrimNm{G}} =
\emptyset$, to give a pre-substitutive family $\relEnv{E}_\Grp$. This
gives us the following lemma and theorem.

@@

Writing $i^n$ for $n$-fold iteration of the group operation, we have:
\begin{theorem}
  Let $m$ and $n$ be natural numbers. For all $n$-by-$m$-matrices of
  natural numbers $A$ and $m$-by-$1$ matricies of integers $b$,
  there exists a quantifier-free program $M$ such that
  \begin{displaymath}
    i_1,\dots,i_m\mathord:\mathsf{G}; \Gamma_\Grp \vdash M : \tyPrim{G}{e_1} \to \dots \to \tyPrim{G}{e_n} \to \tyPrim{G}{e}
  \end{displaymath}
  where $e_j = i_1^{A_{j1}}\cdots i_m^{A_{jm}}$ and $e = i_1^{b_1}\cdots i_m^{b_m}$, iff there is a solution $\vec{x}$
  in the integers to the equation $A \vec x = b$.
\end{theorem}
The uninhabited type $\forall t \mathord: \SynTransl{2}.\
\tyPrim{vec}{t + t} \to \tyPrim{vec}{t}$ from
\exref{ex:uninhabited-type} is an instance of this theorem.  For types
indexed by the scaling group $\SynGL{1}$ the situation is more subtle,
as we have assumed that the type $\tyPrim{real}{s}$ has a polymorphic zero
(see \autoref{fig:real-ops}). We can show, nonetheless, that $\forall
s \mathord: \SynGL{1}.\ \tyPrim{real}{s^2} \to \tyPrim{real}{s}$ is
only inhabited by the constantly zero function by setting
$R^\bullet_{\tyPrimNm{real}} = \{(0,0)\}$. If we augment our sort
$\SynGL{1}$ with a square root operation $s^{1/2}$, then it is
possible to add a primitive square root function and still be able to
prove that the cube root type $\forall s \mathord: \SynGL{1}.\
\tyPrim{real}{s^3} \to \tyPrim{real}{s}$ is only inhabited by the
constantly zero function. This is a consequence of the impossibility
of generating $s$ from $s^3$ with only the abelian group operations
and square root.

>>>>

We define the model of the indexing theory for the two-dimensional
geometry instantiation as follows. Each of the sorts is interpreted just as
its semantic counterpart: $\semSort{\SynTransl{2}} = \Transl{2}$, $
\semSort{\SynGL{2}} = \GL{2}$, $ \semSort{\SynOrth{2}} = \Orth{2}$,
and $ \semSort{\SynGL{1}} = \GL{1}$.
%  \begin{displaymath}
%    \begin{array}{@{}llll}
%      \semSort{\SynTransl{2}} = \Transl{2}
%      &
%      \semSort{\SynGL{2}} = \GL{2}
%      &
%      \semSort{\SynOrth{2}} = \Orth{2}
%      &
%      \semSort{\SynGL{1}} = \GL{1}
%    \end{array}
%  \end{displaymath}
Each of the index operations (e.g.,~the group structure and
determinant) is interpreted by the intended semantic operation, and
clearly satisfy the axioms in \exref{ex:two-dim-geo-axioms}. For the
relational interpretations of $\tyPrim{vec}{B,t}$ and
$\tyPrim{real}{s}$, we use the model-based construction from
\autoref{sec:constr-rel-env}, with $R_{\tyPrimNm{vec}}(B,\vec{t}) = \{
(\vec{v}, B\vec{v} + \vec{t}) \sepbar \vec{v} \in \mathbb{R}^2 \}$ and
$R_{\tyPrimNm{real}}(k) = \{ (x, kx) \sepbar x \in \mathbb{R} \}$. The
types of the primitive vector operations presented in
\autoref{sec:motivating-examples} are collected into a context
$\Gamma_\Geom$ with a straightforward interpretation $\eta_\Geom \in
\ctxtSem{\Gamma_\Geom}$.
  % \begin{displaymath}
  %   \begin{array}{l}
  %     R_{\tyPrimNm{vec}}(B,\vec{t}) = \{ (\vec{v}, B\vec{v} + \vec{t}) \sepbar \vec{v} \in \mathbb{R}^2 \} \\
  %     R_{\tyPrimNm{real}}(k) = \{ (x, kx) \sepbar x \in \mathbb{R} \}
  %   \end{array}
  % \end{displaymath}
%   The definitions of $R_{\tyPrimNm{vec}}$ and $R_{\tyPrimNm{real}}$
%   %are as expected. % given the informal definitions we gave in
% formalise the informal ones from~\autoref{sec:motivating-examples}, and 
%   the definitions of %We have set
%   $R^\bullet_{\tyPrimNm{vec}}$ and $R^\bullet_{\tyPrimNm{real}}$ 
%   %to relate $\vec{0}$ and $0$ to themselves to 
%   account for the
%   polymorphic zero constants we have assumed.
%\end{example}

@@

Consider the type of the triangle area %of a triangle 
function from
\exref{ex:area-of-triangle-1}:
\begin{displaymath}
  \begin{array}{@{}l}
    \mathrm{area} : \forall B\mathord:\SynGL{2}, t\mathord:\SynTransl{2}.\ \\
    \hspace{0.8cm}\tyPrim{vec}{B, t} \to \tyPrim{vec}{B, t} \to \tyPrim{vec}{B, t} \to \tyPrim{real}{\abs{\det B}}
  \end{array}
\end{displaymath}
By \thmref{thm:abstraction}, we can derive the following free theorem. For all $B \in \GL{2}$, $\vec{t} \in \Transl{2}$, and $\vec{x}, \vec{y}, \vec{z} \in \mathbb{R}^2$, we have 
\begin{displaymath}
  \abs{\det B}(\tmSem{\mathrm{area}}\ \vec{x}\ \vec{y}\ \vec{z}) = \tmSem{\mathrm{area}}\ (B\vec{x} + \vec{t})\ (B\vec{y} + \vec{t})\ (B\vec{z} + \vec{t})
\end{displaymath}
Thus, directly from the type of the $\mathrm{area}$ function, we can
see that its index-erasure semantics is (a) invariant under
translations, and (b) if the inputs are subjected to a linear
transformation $B$, the output varies with the absolute value of the
determinant of $B$.

@@

For proving non-definability results, we consider the general case of
a sort $\mathsf{G}$ with the equational theory of abelian groups. We
assume a group-indexed primitive type $\tyPrimNm{G}$ and set of
primitive operations typed by $\Gamma_\Grp$ that include the group
identity $1:\tyPrim{G}{1}$, operation $(\cdot): \forall
a,b\mathord:\mathsf{G}.\ \tyPrim{G}{a} \to \tyPrim{G}{b} \to
\tyPrim{G}{a\cdot b}$ and inverse $\ ^{-1}: \forall
a\mathord:\mathsf{G}.\ \tyPrim{G}{a} \to \tyPrim{G}{a^{-1}}$. The
context $\Gamma_\Grp$ may also include members with types of the form
$\forall \vec{a}\mathord:\mathsf{G}.\ \tyPrim{G}{e_1} \to ...
\tyPrim{G}{e_n} \to \tyPrim{G}{e}$ as long as the result index $e$ is
generated by the set $\{e_1,...,e_n\}$. One instance is the
sort~$\SynTransl{2}$ with $0$, $(+)$ and $\mathrm{negate}$ on
$\tyPrim{vec}{t}$. For the purposes of non-definability results, we
may choose an arbitrary index-erasure semantics that supports the
primitive operations.

We use the construction of definability relational environments
from \autoref{sec:constr-rel-env}, with $R^\bullet_{\tyPrimNm{G}} =
\emptyset$, to give a pre-substitutive family $\relEnv{E}_\Grp$. This
gives us the following lemma and theorem.

@@

Writing $i^n$ for $n$-fold iteration of the group operation, we have:
\begin{theorem}
  Let $m$ and $n$ be natural numbers. For all $n$-by-$m$-matrices of
  natural numbers $A$ and $m$-by-$1$ matricies of integers $\vec{b}$,
  there exists a quantifier-free program $M$ such that:
  \begin{displaymath}
    i_1,\dots,i_m\mathord:\mathsf{G}; \Gamma_\Grp \vdash M : \tyPrim{G}{e_1} \to \dots \to \tyPrim{G}{e_n} \to \tyPrim{G}{e}
  \end{displaymath}
  where $e_j = i_1^{A_{j1}}\cdots i_m^{A_{jm}}$ and $e = i_1^{b_1}\cdots i_m^{b_m}$, iff there is a solution $\vec{x}$
  in the integers to the equation $A \vec x = \vec{b}$.
\end{theorem}
The uninhabited type $\forall t \mathord: \SynTransl{2}.\
\tyPrim{vec}{t + t} \to \tyPrim{vec}{t}$ from
\exref{ex:uninhabited-type} is an instance of this theorem.  For types
indexed by the scaling group $\SynGL{1}$ the situation is more subtle,
as we have assumed that the type $\tyPrim{real}{s}$ has a polymorphic zero
(see \autoref{fig:real-ops}). We can show, nonetheless, that $\forall
s \mathord: \SynGL{1}.\ \tyPrim{real}{s^2} \to \tyPrim{real}{s}$ is
only inhabited by the constantly zero function by setting
$R^\bullet_{\tyPrimNm{real}} = \{(0,0)\}$. If we augment our sort
$\SynGL{1}$ with a square root operation $s^{1/2}$, then it is
possible to add a primitive square root function and still be able to
prove that the cube root type $\forall s \mathord: \SynGL{1}.\
\tyPrim{real}{s^3} \to \tyPrim{real}{s}$ is only inhabited by the
constantly zero function. This is a consequence of the impossibility
of generating $s$ from $s^3$ with only the abelian group operations
and square root.

### 2012-07-10 paper/section-monoid.tex

Above we investigated types indexed by abelian group expressions,
whose elements are taken from some group. We now switch from abelian
groups to commutative monoids.  Our main example is strings indexed by
``taint'' markers: when strings are appended their taint markers are
combined. Thanks to the information in the indices, we are able to
prove an indefinability result (\thmref{thm:monoid-indefinability}).

@@

For the relational environments, we use %make use of 
the compositional
construction in \autoref{sec:constr-rel-env}, setting the chosen model
$\mathcal{M}$ to be the centre $Z(M)$ of $M$. % (recall that the
%centre of a monoid is the set of all elements that commute with every
%other element). 
Setting $R_{\tyPrimNm{M}}(m) = \{ (n,mn) \sepbar n
\in M \}$
%By setting the model $\mathcal{M}$ to be only the elements of $M$ that
%commute with everything else, we ensure that the following lemma
%holds:
gives:
\begin{lemma}\label{lem:monoid-ops-related}
  For all $\Delta$ and $\rho \in \relEnv{E}_\Mon(\Delta)$,
  $(\eta_\Mon,\eta_\Mon) \in \rsem{\Gamma_\Mon}{\relEnv{E}}\rho$.
\end{lemma}

@@

\begin{theorem}
  \label{thm:monoid-indefinability}
  Let $m$, $n$ be natural numbers. For all $n$-by-$m$-matrices of
  natural numbers $A$ and $m$-by-$1$ matrices of natural numbers $b$,
  there is a quantifier-free program $N$, with
  \begin{displaymath}
    i_1,\dots,i_m\mathord:\mathsf{M}; \Gamma_\Mon \vdash N : \tyPrim{M}{e_1} \to \dots \to \tyPrim{M}{e_n} \to \tyPrim{M}{e}
  \end{displaymath}
  where $e_j = i_1^{A_{j1}}\dots i_m^{A_{jm}}$ and $e = i_1^{b_1}\dots
  i_m^{b_m}$ iff there is a solution $\vec{x} = (x_1,...,x_n)$ in the
  natural numbers to $A x^\top = b$.
\end{theorem}

The proof of this theorem uses the non-compositional relational
environments from \autoref{sec:constr-rel-env}, with
$R^\bullet_{\tyPrimNm{M}} = \emptyset$. From the theorem we can see
that it is impossible to write a program with the type
$\tyPrim{M}{ab} \to \tyPrim{M}{a}$, and is thus impossible to
convert a doubly tainted value to a singly tainted one.
%
By contrast with abelian group indexed types, the lack of inverses in
the monoid indexed setting means that the type $\tyPrim{M}{ab} \to
\tyPrim{M}{b} \to \tyPrim{M}{a}$ is also uninhabited since there are
no naturals $x_1,x_2$ such that $(ab)^{x_1}b^{x_2} \equiv a$.

>>>>

Above, we investigated types indexed by abelian group expressions,
whose elements are taken from some group. We now switch from abelian
groups to commutative monoids.  Our main example is strings indexed by
``taint'' markers: when strings are appended their taint markers are
combined. Thanks to the information in the indices, we are able to
prove an indefinability result (\thmref{thm:monoid-indefinability}).

@@

For the relational environments, we use %make use of 
the model-based
construction in \autoref{sec:constr-rel-env}, setting the chosen model
$\mathcal{M}$ to be the centre $Z(M)$ of $M$. % (recall that the
%centre of a monoid is the set of all elements that commute with every
%other element). 
Setting $R_{\tyPrimNm{M}}(m) = \{ (n,mn) \sepbar n
\in M \}$
%By setting the model $\mathcal{M}$ to be only the elements of $M$ that
%commute with everything else, we ensure that the following lemma
%holds:
gives:
\begin{lemma}\label{lem:monoid-ops-related}
  For all $\Delta$ and $\rho \in \relEnv{E}_\Mon(\Delta)$,
  $(\eta_\Mon,\eta_\Mon) \in \rsem{\Gamma_\Mon}{\relEnv{E}}\rho$.
\end{lemma}

@@

\begin{theorem}
  \label{thm:monoid-indefinability}
  Let $m$, $n$ be natural numbers. For all $n$-by-$m$-matrices of
  natural numbers $A$ and $m$-by-$1$ matrices of natural numbers $\vec{b}$,
  there is a quantifier-free program $N$, with:
  \begin{displaymath}
    i_1,\dots,i_m\mathord:\mathsf{M}; \Gamma_\Mon \vdash N : \tyPrim{M}{e_1} \to \dots \to \tyPrim{M}{e_n} \to \tyPrim{M}{e}
  \end{displaymath}
  where $e_j = i_1^{A_{j1}}\dots i_m^{A_{jm}}$ and $e = i_1^{b_1}\dots
  i_m^{b_m}$, iff there is a solution $\vec{x}$ in the
  natural numbers to the equation $A \vec{x} = \vec{b}$.
\end{theorem}

The proof of this theorem uses the definability relational
environments from \autoref{sec:constr-rel-env}, with
$R^\bullet_{\tyPrimNm{M}} = \emptyset$. From the theorem we can see
that it is impossible to write a program with the type
$\tyPrim{M}{ab} \to \tyPrim{M}{a}$, and is thus impossible to
convert a doubly tainted value to a singly tainted one.
%
By contrast with abelian group indexed types, the lack of inverses in
the monoid indexed setting means that the type $\tyPrim{M}{ab} \to
\tyPrim{M}{b} \to \tyPrim{M}{a}$ is also uninhabited since there are
no naturals $x_1,x_2$ such that $(ab)^{x_1}b^{x_2} \equiv a$.
